\chapter{Virtual Memory III, I/O}


How does the processor do I/O?
\begin{enumerate}
	\item Special input/output instructions (e.g.~per device, tuned to hardware).
	\item Memory-mapped I/O: I/O devices are named by memory addresses; load/store using normal instructions. There are a lot of control (key down) and data (what key?) registers.
\end{enumerate}
Two ways to handle I/O:
\begin{description}
	\item[polling] wait for device to set \texttt{ready} bit, then read/write. (\texttt{ready} could be read or write control.)
	\item[interrupt] when I/O is ready or needs attention, control is sent to the \emph{interrupt handler}.
\end{description}
Traps are the software mechanism for dealing with unusual behavior.
\begin{description}
	\item[interrupts] are asynchronous signals (e.g.~I/O is ready) to enter the trap handler.
	\item[exceptions] are synchronous errors (e.g.~illegal memory R/W)
\end{description}
How is an interrupt handled?
\begin{enumerate}
	\item Instruction stream is suspended.
	\item CPU looks up the vector (function address) of interrupt handler in \emph{interrupt vector table} stored in CPU.
	\item \textsc{jal} to the handler, save PC in \textbf{M}achine \textbf{E}xception \textbf{P}rogram \textbf{Counter}.
	\item Handler runs, then returns:
\begin{minted}{asm}
handler: save registers
         decode interrupt
         handle interrupt
         clear interrupt
         restore registers
         mret
\end{minted}
\end{enumerate}
Exceptions are handled like pipeline hazards on a pipelined CPU:
\begin{enumerate}
	\item Complete instructions begun before exception.
	\item Flush instructions currently in pipeline.
	\item (optional) Store exception cause in status register.
\end{enumerate}
