\chapter{Cache Coherence, OpenMP Sharing Issues, Performance}

\section{Multiprocessor Sharing}
Key Questions:
\begin{enumerate}
    \item How do they share data?
    \item How do they coordinate?
    \item How many processors can be supported?
\end{enumerate}

\subsection{Shared Memory Multiprocessor (SMP)}
SMPs contain two or more identical CPUs and a single shared coherent memory. To answer the above questions:`
\begin{enumerate}
    \item Single physical address space shared by all processors/cores
    \item Processors coordinate/communicate through shared variables in memory (via loads and stores)
    \item Depends on the workload
\end{enumerate}

\subsection{Multiprocessor Caches}
\begin{itemize}
    \item Memory is a performance bottleneck even with one processor
    \item Use caches to reduce bandwidth demands on main memory
    \item Each core has a local private cache
    \item Only cache misses have to access the shared common memory
\end{itemize}


\section{Multiprocessor Cache Coherency}
Since there are multiple caches, there might be multiple copies of a piece of data floating around in the system. When any processor has cache miss or writes, we must notify other processors via an interconnection network.

Each cache block tracks the following states:
\begin{description}
	\item[Shared:] up-to-date data, other caches may have a copy.
	\item[Modified:] up-to-date data, changed (dirty), no other cache has a copy, can write back, memory out-of-date.
	\item[Exclusive:] up-to-date data, no other cache has a copy, can write back, memory up-to-date.
	\item[Owner:] up-to-date data, other \emph{shared} caches may have a copy, memory not up-to-date.
\end{description}

\subsection{MOESI}
Memory access to cache is one of the following:
\begin{itemize}
    \item \emph{Modified}: I have the only copy, can write, and it's dirty.
    \item \emph{Owned}: I have the official copy, can write, and it's dirty. When writing, I have to tell everyone else I am writing (and it turns Modified).
    \item \emph{Exclusive}: I have the only copy.
    \item \emph{Shared}: I have a copy and can read away.
    \item \emph{Invalid}
\end{itemize}
\begin{itemize}
    \item Invalid is not in cache
\end{itemize}

\section{Coherency Miss}
If data are shared between two processor cores, every time one does a write, the other will take a \emph{coherency miss} on the next read or write. 